Cache Oblivious B-trees

### 2011-12-09 paper.tex
Fixing the streaming section exception for the insert section

While B-trees achieve good performance for searches, they do not perform as
well for updates and insertions. The buffered-repository tree (BRT) is a
common alternative to the B-tree achieves an improved amortized $O(log(N)/B)$
memory transfers for inserts compared to the $O(log_{B+1}N)$ memory access of
B-trees.

The search vs insertion trade off also exists in the cache-oblivious model.
The original cache-oblivious B-trees described in Section~\ref{sec:original}
achieves the search end of this trade off. However, Bender et al. propose two
data structures that and achieve the opposite end: the shuttle tree and the
cache-oblivious lookahead array (COLA) \cite{BenderFaFi07}. These two data
structures are refereed to as ``streaming B-trees'', indicating how they are
optimized for insertions at the cost of search.

@@

The shuttle tree is similar in layout and structure as the original CO B-Tree.
It uses a strongly weight-balanced search tree described previously (see
Section~\ref{sec:structure}) that is laid out in a vEB layout into a PMA. A
notable difference is that the CO B-tree splits the tree at height $h/2$ when
creating the vEB layout while the shuttle tree splits at height $F_k$ where
$F_k$ is the $kth$ Fibonacci number that is stickily less than the height of
the tree. Thus if a tree's height happens to be a Fibonacci number, $F_k$,
then the shuttle tree will choose to split the tree at height $F_{k-1}$. The
top and bottom ``halves'' produced by this split are organized just like
before and shown in Figure \ref{fig:veb}.

@@

The shuttle tree improves the efficiency of insertion by adding buffers to
each node in the tree. In fact, shuttle tree nodes have a linked list of
buffers rather than a single buffer like BRT and each of these buffers is
another shuttle tree. The length of a node's linked list is related to the
node's height in the tree. A node has a buffer for every time it the leaf of a
sub trees that is split when creating the vEB layout. This idea is illustrated
in Figure \ref{fig:buffers}.

The shuttle tree's node's buffers have a maximum capacity that is enforced by
bounding it height. The size of the buffers increases going across the linked
list. Thus, a node's first buffer is smaller than its last buffer. The maximum
height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil log_{\phi} k \rceil}
$ where $\phi$ is the golden ratio and $k$ is chosen such that $F_k$ =
$\xi(h)$ and $\xi(x)$ is defined recursively. If $h$ is a Fibonacci number
then $\xi(h) = h$ and otherwise $\xi(h) = \xi(h-f)$ where f is the largest
Fibonacci number less than $h$. The size of the buffer is related to the size
of the recursive subtree for which the node is a leaf. Since the recursive
subtrees are split based on the largest Fibonacci number less than $h$ it
follows that the buffer sizes are also based on Fibonacci numbers.

@@

The operations that are described by Bender et al. for the shuttle tree are
\Search, \Insert, and \Scan. An explicit discussion on \Delete is missing;
however, it is assumed that it is symmetric to \Insert.

\Search and \Scan work virtually the same way as the CO B-tree. The only
difference is that you have to search a child node's buffer before following
the pointer down the tree.

\Insert is also similar to the CO B-tree except you buffer elements as you add
new elments. When you insert an element you move down the tree from the root
to leaves and stop when you find a node that has a buffer (it is possible that
a node does not have any buffers because of the height of the tree). You then
attempt to insert the element into the smallest buffer in the node's linked
list. If this insertion causes the buffer to overflow (e.g. the height of the
buffer exceeds its maximum), you move all of the elements from this buffer to
the next largest buffer. When the largest buffer overflows you take all of its
elements and ``shuttle'' them down the tree. Eventually you will hit a leaf
node and inserting here may cause the balance constraints on the SWBST to
become invalidated. The shuttle tree rebalances itself just like any other
SWBST: the leaf node is split into two and this cascades up the tree until
balance is restored.

When a node $u$ is split into two nodes $u1$ and $u2$ we immediately allocate
enough space for all of the $u2$'s buffers. We then scan over $u$'s buffered
sub trees and find the ones rooted in $u1$ and $u2$ and move them
respectively. This can be achieved by a single scan. Once we have found the
set of subtrees that correspond to $u1$ and $u2$ we replace the original sub
trees for $u$ in the vEB layout.

The core problem of maintain room for inserts in the shuttle tree does not
change from the CO B-tree and is handled by the PMA just like in described in
Section~\ref{sec:layout}.

@@

The complexity of \Search is $O(log_B N)$ where $N = \Theta(c^{F_k})$ and $c$
is your fanout. Note that $c$ is not treated as a constant since this is a
cache-oblivious data structure. This bound is achieved by searching over the
top and bottom recursive subtrees at each level. A key trick is that when you
visit a node you only search its largest buffer (that last buffer in the
linked list). The smaller buffers will be visited eventually as you recurse
down the tree. Eventually you will get down to a single subtree that can fit
into $O(1)$ blocks and thus can be search with $O(1)$ memory accesses. Since
the fanout is proportional to our block size it will take us $O(log_B N)$
block accesses until we get down to a subtree and thus, by induction, the
worst case runtime is $O(log_B N)$.

@@

The complexity to \Scan $L$ elements is amortized to $O(L/B)$ memory
transfers. First you perform a \Search to get to the starting element of your
range. Once you have a block in memory you can start scanning in elements and
since you only have to read a new block every $B$ accesses each element access
is amortized to $1/B$ and thus overall the number of memory access is
$O(L/B)$. This, of course, assumes that you just performed a search for the
starting point of your scan.

The last thing that should be considered is the space requirement of the
shuttle tree. However, the overall space requirements of the shuttle tree is
actually just $O(n)$. The high level intuition for this proof is that the
majority of the buffers are really small and thus do not impact the overall
size of the tree.

@@

The cache-oblivious lookahead array (COLA) is a cache-oblivious dictionary
that is actually quite different from the other cache-oblivious data
structures seen so far. Instead of an explicit tree structure with child and
parent pointers, the COLA is organized as a list of arrays or levels. Each
level has arrays of size $2^k$. Each level is either completely full or
completely empty. As such, the COLA is very similar to the binomial list
structure described in \cite{BentleySaxe}.

@@

The operations supported by COLA are \Search, \Insert, and \Scan.

\Insert works by initially inserting the element into the first-level array.
This array has size $2^0 = 1$, thus the element is in an array on it own.
Whenever you insert an element you check to see if there is another list of
the same size and merge them together in sorted order. This merging will
produce a new array of twice the size which goes into the next level. We
continue to merge until there are no two arrays of the same length.

\Search can naively be performed by simply scanning each of the levels in our
data structure. Since everything is sorted we can use a binary search to
quickly find the element we are looking for. Obviously this is not very
efficient, thus COLA, as the name suggest, uses look ahead pointers to reduce
the number of memory accesses required. Every 8th element from the $k+1$ level
exists in the $k$ level array in its proper sorted location. This element also
contains a pointer to its real location in the $k+1$ level, thus it is
referred to as a ``real lookahead pointer''. Additionally, each every 4th cell
in $k$ level array contains two pointers pointing to the next and previous
``real lookahead pointers'' on this level. Therefore, to search a level we
perform a binary search and automatically cascade up to the next level.

@@

The complexity of \Insert is amortized to $O(log N / B)$ block transfers.
First, it is important to know that an element will never be merged more than
$O(log(N))$ times since at that point, it will be in the last array. The first
$log(B)$ merges that a element participates is constant time since we can
assume that the relevant blocks are already in memory. Thus, we only need to
worry about memory accesses when we merge levels that have size greater than
$B$. If the size of the array is $k$ then the cost of merging is $O(k/B)$ and
thus the amortized cost of a merging a single element is simply $1/B$. Since
an element will only be merged $log(N)$ time, the overall cost is amortized to
$O(logN/B)$.

The complexity of \Search is $O(log N)$ memory transfers. The proof is based
on the fact that we only ever need to search at most eight items on each
level. Since there are $O(log(N))$ levels and since eight is a constant it
follows that overall, search accesses $O(log N)$ blocks. A proof by induction
shows that only eight items need to be searched on each level. The base case
is trivial. The first three levels of the COLA each have at most eight
elements that either contain the actual elements or a look ahead pointer to
the next level. Obviously we can search these levels with no more than eight
member transfers. As we follow the lookahead pointers to the high levels we
will eventual get to a level where $k \geq 3$. Note that we do not enter this
level at the start; rather, we arrived here by following the real lookahead
pointers from the levels below and are at some index $i$ that contains an
element $e_i$ such that $e_i < e$ where $e$ is the element we are looking for.
Our goal is to find a new $i$ such that $e_i \leq e < e_{i+1}$. Once we find
this $i$ then we can use the duplicate look ahead pointer to find the ``next''
and previous look ahead pointers in the next level and restrict our search to
that range. Since we have a real look ahead pointer for every eighth element
in the next level, it follow that when we get to the next level, we only need
to search eight elements. Furthermore, since we arrived at our current level
from a lower level, we only need to search $8$ elements to find this $i$.
Thus, overall, we only need to perform $O(8log(N)) = O(log(N))$ memory
transfers.

@@

In experimental results, COLA achieved significant improvements for random
inserts, performing 790 times faster than a normal B-tree. Sorted inserts are
3.1 times slower and searches are 3.5 times slower. These experimental results
indicate that the the COLA is a suitable data structure for streaming data
structure where random insertions are common.

>>>>

While B-trees achieve an optimal bound for the number of memory transfer for
search, they do not perform as well for updates and insertions. The
buffered-repository tree (BRT) is a common alternative to the B-tree that
achieves an improved amortized $O(log(N)/B)$ memory transfers for inserts
compared to the $O(log_{B+1}N)$ memory access of B-trees. Of course, the
B-tree and BRT are analyzed with the DMA model and are not cache-oblivious.

This search vs insertion trade off also exists in the cache-oblivious model.
The original BO B-tree described in Section~\ref{sec:original} is designed to
optimize for the search in this trade off. However, Bender et al. propose two
data structures that and achieve the opposite end: the shuttle tree and the
cache-oblivious lookahead array (COLA) \cite{BenderFaFi07}. These two data
structures are refereed to as ``streaming B-trees'', indicating how they are
optimized for insertions at the cost of search.

@@

The shuttle tree is similar in layout and structure as the original CO B-Tree.
It uses a strongly weight-balanced search tree described previously (see
Section~\ref{sec:structure}) that is laid out in a vEB (see
Section~\ref{sec:veb}) layout into a PMA (see Section~\ref{sec:layout}). A
notable difference between the shuttle tree and the CO B-tree is that the CO
B-tree splits the tree at height $h/2$ when creating the vEB layout while the
shuttle tree splits at height $F_k$, where $F_k$ is the $kth$ Fibonacci number
that is stickily less than the height of the tree. Thus if a tree's height
happens to be a Fibonacci number, $F_k$, then the shuttle tree will choose to
split the tree at height $F_{k-1}$. The top and bottom ``halves'' produced by
this split are organized just like before and shown in Figure \ref{fig:veb}.

@@

At a high level, the shuttle tree improves the efficiency of insertion by
adding buffers to each node in the tree. In fact, shuttle tree nodes have a
linked list of buffers rather than just a single buffer like the BRT. Each of
these buffers is another shuttle tree. The a shuttle tree node enforces two
constraints: the length of its linked list as well as the overall height (and
thus capacity) of each of the buffered shuttle trees contained within its
linked list.

The length of a node's linked list is related to the node's height in the
tree. A node has a buffer for every time it the leaf of a sub trees that is
split when creating the vEB layout. This idea is illustrated in Figure
\ref{fig:buffers}.

The shuttle tree's node's buffers have a maximum capacity that is enforced by
bounding its height. The size of the buffers increases going across the linked
list. Thus, a node's first buffer is smaller than its last buffer. The maximum
height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil log_{\phi} k \rceil}
$ where $\phi$ is the golden ratio and $k$ is chosen such that $F_k$ =
$\xi(h)$ and $\xi(x)$ is defined recursively. If $h$ is a Fibonacci number
then $\xi(h) = h$ and otherwise $\xi(h) = \xi(h-f)$ where f is the largest
Fibonacci number less than $h$. The maximum height of an individual buffer is
related to the size of the recursive subtree for which the node is a leaf.
Since the recursive subtrees are split based on the largest Fibonacci number
less than $h$ (as described previously) it follows that the buffer sizes are
also based on Fibonacci numbers.

@@

The operations that are described by Bender et al. for the shuttle tree are
\Search{}, \Insert{}, and \Scan{}. An explicit discussion on \Delete{} is
missing; however, it is assumed that it is symmetric to \Insert{}.

\Search{} and \Scan{} work virtually the same way as the CO B-tree. The only
difference is that you have to search a child node's buffer before following
the pointer down the tree.

\Insert{} is also similar to the CO B-tree except you buffer elements as you
add them. When you insert an element you move down the tree from the root to
leaves and stop when you find a node that has a buffer (note that because of
the nature of Fibonacci numbers that it is possible that a node does not have
any buffers). You then attempt to insert the element into the smallest buffer
in the node's linked list. If this insertion causes the buffer to overflow
(e.g. the height of the buffer exceeds its maximum), you move all of the
elements from this buffer to the next largest buffer. When the largest buffer
overflows you take all of its elements and ``shuttle'' them down the tree.
Eventually you will hit a leaf node and inserting here may cause the balance
constraints on the SWBST to become invalidated. The shuttle tree rebalances
itself just like any other SWBST: the leaf node is split into two and this
split cascades up the tree until balance is restored.

The shuttle tree also needs to take care of the buffers belonging to a split
node. When a node $u$ is split into two nodes $u_1$ and $u_2$ the shuttle tree
immediately allocates enough space for all of the $u_2$'s buffers ($u_1$
simply reuses the space allocated for $u$'s buffers). We then scan over $u$'s
buffers and find the ones rooted in $u_1$ and $u_2$ and move them
respectively. This can be achieved by a single scan.

The core problem of maintaining room for inserts, bounding the costs of
merging and splitting, and bounding the costs of updating parent and child
pointers in the shuttle tree is achieved using the same PMA used with the CO
B-tree. This is decried in detail in Section~\ref{sec:layout}.

@@

\Search uses $O(log_B N)$ memory transfers where $N = \Theta(c^{F_k})$ and $c$
is your fanout. Note that $c$ is not treated as a constant since this is a
cache-oblivious data structure. This bound is achieved by searching over the
top and bottom recursive subtrees at each level when traversing the tree. A
key trick is that when you visit a node you only search its largest buffer
(that last buffer in the linked list for that level). The smaller buffers will
be visited eventually as you recurse down the tree. Eventually you will get
down to a single subtree that can fit into $O(1)$ blocks and thus can be
searched with $O(1)$ memory accesses. Since the fanout is proportional to our
block size it will take us $O(log_B N)$ block accesses until we get down to a
subtree and thus, by induction, the worst case runtime is $O(log_B N)$.

*****

@@

******

\Scan{} uses an amortized $O(L/B)$ memory transfers to scan in $L$ consecutive
elements. This cost obviously does not include the single \Search{} operation
that is required to find the first starting element of the range. Once that
first element's block is transferred, the shuttle tree proceeds by traversing
the tree and scanning in elements. Since the vEB layout enforces the spatial
locality of nearby nodes in the tree, the shuttle tree only has to read in a
new block once every $B$ memory accesses. Thus, each individual memory access
is is amortized to $1/B$ and the overall number of memory transfers is
$O(L/B)$.

The last thing that is worth mentioning is that the overall space requirements
of the shuttle tree is $O(n)$. The high level intuition for this proof is that
the majority of the buffers are really small and thus do not impact the
overall size of the tree.

@@

The cache-oblivious lookahead array (COLA) is actually quite different from
the other cache-oblivious data structures seen so far. Instead of an explicit
tree structure with child and parent pointers, the COLA is organized as a list
of arrays or levels. Each array has size $2^k$. Furthermore, each array is
either completely full or completely empty. As such, the COLA is very similar
to the binomial list structure described in \cite{BentleySaxe}.

@@

The operations supported by COLA are \Search{}, \Insert{}, and \Scan{}. Once
again, \Delete{} can be inferred symmetrically from \Insert{}.

\Insert{} works by initially inserting an element into the first-level array.
This array has size $2^0 = 1$, thus the element is in an array on it own.
Whenever you insert an element you check to see if there is another array of
the same size (this can be though of as checking to see if there is another
array in the same level). If there is another array with the same size, the
COLA merges the two together in sorted order. This merging will produce a new
array that is twice the size and thus goes into the next level. The COLA
continues to merge until there are no two arrays of the same size.

\Search{} can naively be performed by simply scanning each of the levels in
the COLA. Since everything is sorted the COLA could can use a binary search to
quickly find the element it is looking for. However, this is not very
efficient. As an improvement COLA, as the name suggest, uses look ahead
pointers to reduce the number of memory accesses required. Every 8th element
from the $(k+1)$st level exists in the $k$ level array in its proper sorted
location. This element also contains a pointer to its real location in the
$(k+1)$st level. Thus this pointer is referred to as ``real lookahead
pointer''. Additionally, each every 4th cell in $k$th level array contains two
pointers pointing to the next and previous ``real lookahead pointers'' on this
level. The next and previous pointers are referred to as ``duplicate lookahead
pointers''. To search a level the COLA performs a binary search within an
appropriate range, determine the range in the next level using the ``duplicate
lookahead pointers'' and then recursively cascade up the tree.

@@

\Insert{} achieves an amortized $O(log N / B)$ memory transfers. The proof is
based on the far that an element will never be merged more than $O(log(N))$
times. This is because at that point it will be in the last array.
Additionally, the first $log(B)$ merges that a element participates in uses a
constant number of memory transfer since we can assume that the relevant
blocks are already cached. Thus, we only need to worry about memory transfers
when we merge levels that have size greater than $B$. If the size of the array
is $k$ then the cost of merging is $O(k/B)$ and thus the amortized cost of a
merging a single element is simply $1/B$. Since we just proved that an element
will only be merged $O(log(N))$ time, it follows that the overall amortized
memory transfers to insert an element is $O((log N)/B)$.

\Search{} achieves $O(log N)$ memory transfers. The proof uses induction to
show that the COLA never needs to search more than eight elements on each
level. Since there are $O(log(N))$ levels and since eight is a constant, it
follows that overall, search uses $O(log N)$ memory transfers. The fact that
only eight elements need to be search per level is shown through induction.
The base case is trivial. The first three levels of the COLA each have at most
eight elements (note that these elements are either actual real elements or
lookahead pointers). The COLA can obviously search these levels with no more
than eight member transfers. As the COLA follows lookahead pointers to the
higher levels, it will eventual get to a level where $k \geq 3$. Note that the
COLA does not ``enter'' this level at the start of the array; rather, its
arrived here by following the real lookahead pointers from the levels below
and thus is at some index $i$ that contains an element $e_i$ such that $e_i <
e$ and $e$ is the element being searched for. The goal at this point is to
find a new $i$ such that $e_i \leq e < e_{i+1}$. Once this $i$ is found the
duplicate lookahead pointers can be used to locate the ``next'' and
``previous'' real lookahead pointers in the next level. The two real lookahead
pointers are used to and restrict the search in the next level. Recall that at
each level in the COLA every eighth element in a level is present in the
previous level. Thus, it follows that in next level only 8 elements need to be
searched. Furthermore, since the current level was reach from a lower level,
it follows that only 8 elements need to be search on this level to find the
appropriate $i$. Thus, overall, the COLA uses $O(8log(N)) = O(log(N))$ memory
transfers.

@@

In experimental results, the COLA achieved significant improvements over
normal (cache-aware) B-trees, performing 790 times faster. However this
improvement comes at the cost of sorted inserts and normal searches. Sorted
inserts are 3.1 times slower and searches are 3.5 times slower. Nevertheless,
these experimental results indicate that the the COLA is a suitable data
structure when random insertions are common.
